The \dealii Library, Version 9.2

### 2020-05-19 9.2/paper.tex
Update.

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
%     Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
    Peter Munch,
%     Ross Maguire Kynch,
    Matthias Maier,
%     Jean-Paul Pelteret,
%     Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

% \author[3]{Thomas~C.~Clevenger}
% \affil[3]{School of Mathematical and Statistical Sciences,
%   Clemson University,
%   Clemson, SC, 29634, USA
%   {\texttt{\{tcleven, heister\}@clemson.edu}}}
%
% \author[4]{Denis~Davydov}
% \affil[4]{Chair of Applied Mechanics,
%   Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
%   Egerlandstr.\ 5,
%   91058 Erlangen, Germany.
%   {\texttt{\{denis.davydov,jean-paul.pelteret\}@fau.de}}}
%
% \author[5]{Marc~Fehling}
% \affil[5]{Institute for Advanced Simulation,
%   Forschungszentrum J\"{u}lich GmbH,
%   52425 J\"{u}lich, Germany.
%   {\texttt{m.fehling@fz-juelich.de}}}
%
% \author[6]{Daniel Garcia-Sanchez}
% \affil[6]{Sorbonne Universit\'es, UPMC Univ.\ Paris 06, CNRS-UMR 7588,
%   Institut des NanoSciences de Paris, F-75005, Paris, France
%   {\texttt{daniel.garcia-sanchez@insp.upmc.fr}}}
%
% \author[2]{Graham Harper}
%
\author[3]{Timo~Heister}
 \affil[3]{School of Mathematical and Statistical Sciences,
   Clemson University,
   Clemson, SC, 29634, USA
   {\texttt{heister@clemson.edu}}}

@@

%
% \author[4]{Jean-Paul~Pelteret}
%
% \author[*1]{Bruno~Turcksin}
%
% \author[12]{David Wells}
% \affil[12]{Department of Mathematics, University of North Carolina,
%   Chapel Hill, NC 27516, USA.
%   {\texttt{drwells@email.unc.edu}}}

@@

The major changes of this release are:
%
\begin{itemize}
\item xy \todo[inline]{Update once we have the subsections in Section 2;
  provide cross-references to each of these subsections}
%   \item Improved support for automatic differentiation (see
%     Section~\ref{subsec:ad}), 
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item x \todo[inline]{Wolfgang to write about complex-valued output}
\item y \todo[inline]{Timo to write about problems and fixes for
  computations with more than $2^{32}$ unknowns}
\item z \todo[inline]{What else? Maybe mention the updated step-12?}
\end{itemize}
%
In addition to these changes, the changelog lists more than 200 other
features and bugfixes.

@@

By release 9.1, all triangulation classes of \texttt{deal.II} have in common that the coarse grid is shared by 
all processes and the actual mesh used for computations is constructed by repeated 
refinement. However, this has its limitations in industrial applications where, often, the mesh comes 
from an external mesh generator in the form of a file that already contains millions 
or tens of millions of cells. For such configurations, applications might already 
run out of  main memory, while reading the complete mesh by each MPI process.

The new class \texttt{parallel::fullydistributed::Triangulation} targets this issue 
by distributing also the coarse grid. Such 
a triangulation can be created by providing to each process a \texttt{Triangulation\-De\-scrip\-tion::Description} struct, containing 
1) the relevant data to construct the local part of the coarse grid, 2) the 
translation of the local coarse-cell IDs to globally unique IDs, 3) the hierarchy 
of mesh refinements, and 4) the owner of the cells on the active mesh level as well 
as on the multigrid levels. Once the triangulation is set up with this struct, no adaptive
changes to the mesh are allowed at the moment.

@@

The new fully distributed triangulation class works---in contrast to  
\texttt{parallel::distributed::\allowbreak Tri\-an\-gu\-la\-tion}---not only for 2D- and 3D- but also for 
1D-problems. It can be used in the context of geometric multigrid methods and 
supports periodic boundary conditions. Furthermore, hanging nodes are supported.

@@

The class \texttt{VectorizedArray<Number>} is a key ingredient for the high 
node-level performance of the matrix-free algorithms in deal.II~\cite{KronbichlerKormann2012, KronbichlerKormann2019}. It is a wrapper 
class around $n$ vector entries of type \texttt{Number} and delegates relevant 
function calls to appropriate Intrinsics instructions. Up to release 9.1, the 
vector length $n$ has been set at compile time of the library to the highest 
possible value supported by the given processor architecture.

The class \texttt{VectorizedArray} has been made more user-friendly by making 
it compatible with the STL algorithms found in the header \texttt{<algorithm>}. 
The length of the  vector can now be queried by  \texttt{VectorizedArray::size()} and its underlying number type by \texttt{VectorizedArray::value\_type}. 
Furthermore, the \texttt{VectorizedArray} class supports range-based iterations over its entries and, i.a., the following 
algorithms: \texttt{std::\allowbreak ad\-vance()}, \texttt{std::distance()}, and \texttt{std::max\_element()}.

This class has been also extended with the second optional template argument 
\texttt{VectorizedArray<Number, size>} with \texttt{size} being related to the 
vector length, i.e., the number of lanes to be used and the instruction set to be 
used. By default, the number is set to the highest value supported by the given 
hardware. A full list of supported 
vector lengths is presented in Table~\ref{tab:vectorizedarray}.

All matrix-free related classes (like \texttt{MatrixFree} and \texttt{FEEvaluation}) 
have been templated with the floating-point number type \texttt{Number} (e.g. \texttt{double} or \texttt{float}); the computations were performed implicitly 
on \texttt{VectorizedArray<Number>} structs with the highest 
instruction-set-architecture extension, with each lane responsible for a separate 
cell (vectoriziation over elements). In release 9.2, all matrix-free classes 
have been extended with a new optional template argument specifying the 
\texttt{VectorizedArrayType}. This allows users to select the vector length/ISA and, 
as a consequence, the number of cells to be processed at once directly in their applications: 
The deal.II-based 
library \texttt{hyper.deal}~\cite{munch2020hyperdeal}, which solves the 6D Vlasov--Poisson equation with high-order 
discontinuous Galerkin methods (with more than 1024 degrees of freedom per cell), works with \texttt{MatrixFree} objects of different SIMD-vector 
length in the same application and benefits---in terms of performance---by the possibility of decreasing the number of cells processed by a single SIMD instruction.

\begin{table}
\caption{Supported vector lengths of the class \texttt{VectorizedArray} and 
the corresponding instruction-set-architecture extensions. }\label{tab:vectorizedarray}
\centering
\begin{tabular}{ccc}
\toprule
\textbf{double} & \textbf{float} & \textbf{ISA}\\
\midrule
VectorizedArray<double, 1> & VectorizedArray<float, 1> & (auto-vectorization) \\
VectorizedArray<double, 2> & VectorizedArray<float, 4> & SSE2 \\ 
VectorizedArray<double, 4> & VectorizedArray<float, 8> & AVX/AVX2 \\ 
VectorizedArray<double, 8> & VectorizedArray<float, 16> & AVX-512 \\ 
\bottomrule
\end{tabular}

\caption{Comparison of relevant SIMD-related classes in deal.II and C++20.}\label{tab:simd}
\centering
\begin{tabular}{cc}
\toprule
\textbf{VectorizedArray (deal.II)} & \textbf{std::simd (C++20)} \\
\midrule
VectorizedArray<Number> & std::experimental::native\_simd<Number> \\
VectorizedArray<Number, size> & std::experimental::fixed\_size\_simd<Number, size> \\ \bottomrule
\end{tabular}
\end{table}

A side effect of introducing the new template argument \texttt{VectorizedArrayType} 
in the \texttt{MatrixFree} classes is that any data structures 
\texttt{VectorizedArrayType} can be processed if they support required 
functionalities like \texttt{size()} or \texttt{value\_type}. In this context, we 
would like to highlight that the new \texttt{C++20} feature \texttt{std::simd} 
can be processed by the matrix-free infrastructure with minor internal 
adjustment as an open pull request shows 
(see \url{https://github.com/dealii/dealii/pull/9994}).  
Table~\ref{tab:simd} gives a comparison of the deal.II-specific SIMD classes and 
the equivalent C++20 classes. 
%We welcome the standardization of the SIMD 
%parallelization paradigm in C++ and intend to replace step by step our own 
%wrapper class, which has been continuously developed over the last decade. We 
%would like to emphasize that the work invested in this class was not in vain, 
%since many performance-relevant utility functions implemented with \texttt{VectorizedArray} in mind (e.g., \texttt{vectorized\_load\_and\_transpose} 
%and \texttt{vectorized\_transpose\_and\_store}) will be still used, since they 
%have not become part of the standard.

@@

In a next step, we intend to support that users could set 
\texttt{VectorizedArrayType} to \texttt{Number}, which would lead to the usage not 
of \texttt{VectorizedArray<Number, 1>} but of a specialized code-path exploiting 
vectorization within an element~\cite{KronbichlerKormann2019}.

@@

Large-scale simulations with 304,128 cores have revealed bottlenecks in release 
9.1 during initialization due to the usage of expensive collective operations 
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak Alltoall()}, e.g., during the 
pre-computation of the index ranges of all processes, which have been stored in  an array. 
This information is needed to set up 
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class. 
In release 9.2, we have removed such arrays and have replaced 
the \texttt{MPI\_Allgather}/\allowbreak\texttt{MPI\_\allowbreak Alltoall} 
function calls by consensus algorithms~\cite{hoefler2010scalable}, which can be 
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}): now, only the locally relevant information about the index ranges is (re)computed when needed, using these algorithms. 
We provide two flavors of the consensus algorithm: the two-step 
approach \texttt{CA::PEX} and the \texttt{CA::NBX}, 
which uses only point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

By replacing the collective communications during setup and removing the arrays 
that contain information for each process (enabled by the application of consensus 
algorithms and other modifications---a full list of modifications leading to this 
improvement can be found online), we were able to significantly improve the setup 
time for large-scale simulations and to solve a Poisson problem with multigrid 
with 12T unknowns. 
Figure~\ref{} compares the timings of a simulation (incl. setup) with the 
previous release 9.1 and with the current release 9.2. 
{\color{red}TODO[Peter/Martin] description of the results}

The new code has been also applied to solve problems with adaptively refined 
meshes with more than 4B unknowns. {\color{red}TODO[Timo]}

@@

\todo[inline]{Marc: Your section}

@@

@@

\item \texttt{step-65} presents \texttt{TransfiniteInterpolationManifold}, a
manifold class that can propagate curved boundary information into the
interior of a computational domain, and \texttt{MappingQCache}, which can sample 
the information of expensive manifolds in the points of a \texttt{MappingQ} and 
cache it for further use.

\item \texttt{step-67} is an explicit time integrator for the
compressible Euler equations discretized with a high-order discontinuous
Galerkin scheme using the matrix-free infrastructure. Besides the use of
matrix-free evaluators for systems of equations and over-integration, it also
presents \texttt{MatrixFreeOperators::CellwiseInverseMassMatrix}, a fast implementation
of the action of the inverse mass matrix in the DG setting using tensor
products. Furthermore, this tutorial demonstrates the usage of new 
pre and post operations, which can be passed to \texttt{cell\_loop()}, to schedule operations on sections of vectors close 
to the matrix-vector product to increase data locality 
and discusses performance-related aspects.

@@

\item \texttt{step-70}
  \todo[inline]{Also need to update announce and announce-short if this
    makes it into the release.}
\end{itemize}

@@

The original \texttt{\dealii} paper containing an overview of its
architecture is \cite{BangerthHartmannKanschat2007}. If you rely on
specific features of the library, please consider citing any of the
following:
\begin{itemize}
 \item For geometric multigrid: \cite{Kanschat2004,JanssenKanschat2011,ClevengerHeisterKanschatKronbichler2019};
 \item For distributed parallel computing: \cite{BangerthBursteddeHeisterKronbichler11};
 \item For $hp$~adaptivity: \cite{BangerthKayserHerold2007};
  \item For partition-of-unity (PUM) and enrichment methods of the
    finite element space: \cite{Davydov2016};
 \item For matrix-free and fast assembly techniques:
   \cite{KronbichlerKormann2012,KronbichlerKormann2019};
 \item For computations on lower-dimensional manifolds:
   \cite{DeSimoneHeltaiManigrasso2009};
 \item For integration with CAD files and tools:
   \cite{HeltaiMola2015};
 \item For Boundary Elements Computations:
   \cite{GiulianiMolaHeltai-2018-a};
 \item For \texttt{LinearOperator} and \texttt{PackagedOperation} facilities:
   \cite{MaierBardelloniHeltai-2016-a,MaierBardelloniHeltai-2016-b}.
 \item For uses of the \texttt{WorkStream} interface:
   \cite{TKB16};
   \item For uses of the \texttt{ParameterAcceptor} concept, the
     \texttt{MeshWorker::ScratchData} base class, and the
     \texttt{ParsedConvergenceTable} class: \cite{SartoriGiulianiBardelloni-2018-a}.
\end{itemize}

\dealii can interface with many other libraries:
\begin{multicols}{3}
\begin{itemize}
\item ADOL-C \cite{Griewank1996a,adol-c}
\item ARPACK \cite{arpack}
\item Assimp \cite{assimp}
\item BLAS and LAPACK \cite{lapack}
\item cuSOLVER \cite{cusolver}
\item cuSPARSE \cite{cusparse}
\item Gmsh \cite{geuzaine2009gmsh}
\item GSL \cite{gsl2016}
\item Ginkgo \cite{ginkgo-web-page}
\item HDF5 \cite{hdf5}
\item METIS \cite{karypis1998fast}
\item MUMPS \cite{ADE00,MUMPS:1,MUMPS:2,mumps-web-page}
\item muparser \cite{muparser-web-page}
\item nanoflann \cite{nanoflann}
\item NetCDF \cite{rew1990netcdf}
\item OpenCASCADE \cite{opencascade-web-page}
\item p4est \cite{p4est}
\item PETSc \cite{petsc-user-ref,petsc-web-page}
\item ROL \cite{ridzal2014rapid}
\item ScaLAPACK \cite{slug}
\item SLEPc \cite{Hernandez:2005:SSF}
\item SUNDIALS \cite{sundials}
\item SymEngine \cite{symengine-web-page}
\item TBB \cite{Rei07}
\item Trilinos \cite{trilinos,trilinos-web-page}
\item UMFPACK \cite{umfpack}
\end{itemize}
\end{multicols}
Please consider citing the appropriate references if you use interfaces to these
libraries.

@@

D.~Arndt and M.~Kronbichler were partially supported by the German
Research Foundation (DFG) under the project ``High-order discontinuous
Galerkin for the exa-scale'' (\mbox{ExaDG}) within the priority program ``Software
for Exascale Computing'' (SPPEXA).

@@

M.~Kronbichler was also supported by the Bayerisches Kompetenznetzwerk
f\"ur Technisch-Wissen\-schaft\-li\-ches Hoch- und H\"ochstleistungsrechnen
(KONWIHR) in the context of the project
``Performance tuning of high-order discontinuous Galerkin solvers for
SuperMUC-NG''.

>>>>

\usepackage{siunitx}

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
%     Thomas C. Clevenger,
%     Denis Davydov,
    Marc Fehling,
%     Daniel Garcia-Sanchez,
%     Graham Harper,
    Timo Heister,
%     Luca Heltai,
    Martin Kronbichler,
    Peter Munch,
%     Ross Maguire Kynch,
    Matthias Maier,
%     Jean-Paul Pelteret,
    Bruno Turcksin,
%     David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

@@

% \author[3]{Thomas~C.~Clevenger}
% \affil[3]{School of Mathematical and Statistical Sciences,
%   Clemson University,
%   Clemson, SC, 29634, USA
%   {\texttt{\{tcleven, heister\}@clemson.edu}}}
%
% \author[4]{Denis~Davydov}
% \affil[4]{Chair of Applied Mechanics,
%   Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
%   Egerlandstr.\ 5,
%   91058 Erlangen, Germany.
%   {\texttt{\{denis.davydov,jean-paul.pelteret\}@fau.de}}}
%
\author[5]{Marc~Fehling}
\affil[5]{Institute for Advanced Simulation,
  Forschungszentrum J\"{u}lich GmbH,
  52425 J\"{u}lich, Germany.
  {\texttt{m.fehling@fz-juelich.de}}}
%
% \author[6]{Daniel Garcia-Sanchez}
% \affil[6]{Sorbonne Universit\'es, UPMC Univ.\ Paris 06, CNRS-UMR 7588,
%   Institut des NanoSciences de Paris, F-75005, Paris, France
%   {\texttt{daniel.garcia-sanchez@insp.upmc.fr}}}
%
% \author[2]{Graham Harper}
%
\author[3]{Timo~Heister}
 \affil[3]{School of Mathematical and Statistical Sciences,
   Clemson University,
   Clemson, SC, 29634, USA
   {\texttt{heister@clemson.edu}}}

@@

%
% \author[4]{Jean-Paul~Pelteret}
%
\author[*1]{Bruno~Turcksin}
%
% \author[12]{David Wells}
% \affil[12]{Department of Mathematics, University of North Carolina,
%   Chapel Hill, NC 27516, USA.
%   {\texttt{drwells@email.unc.edu}}}

@@

The major changes of this release are:
%
\begin{itemize}
\item xy \todo[inline]{Update once we have the subsections in Section 2;
  provide cross-references to each of these subsections}
%   \item Improved support for automatic differentiation (see
%     Section~\ref{subsec:ad}),
\end{itemize}
%
These major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item \dealii{} had decent support for solving complex-valued problems
  (e.g., ones in quantum mechanics -- like the example covered by the
  \texttt{step-58} tutorial program covered below -- or for
  time-harmonic problems) for a while already. However, there were two
  areas in which support was missing. First, the UMFPACK direct solver
  packaged with \dealii{} did not support solving complex-valued
  linear problems. This has now been addressed: UMFPACK actually can
  solve such systems, we just needed to write the appropriate
  interfaces. Second, the \texttt{DataOut} class that is responsible
  for converting nodal data into information that can then be written
  into files for visualization did not know how to deal with vector-
  and tensor-valued fields whose components are complex numbers. An
  example for this is to solve the time-harmonic version of the
  Maxwell equations that has the electric and magnetic fields as
  solution. This, too, has been addressed in this release.
\item z \todo[inline]{What else? Maybe mention the updated step-12?}
\end{itemize}
%
In addition to these changes, the changelog lists more than 200 other
features and bugfixes.

@@

By release 9.1, all triangulation classes of \texttt{deal.II} have in common that the coarse grid is shared by
all processes and the actual mesh used for computations is constructed by repeated
refinement. However, this has its limitations in industrial applications where, often, the mesh comes
from an external mesh generator in the form of a file that already contains millions
or tens of millions of cells. For such configurations, applications might exhaust
available memory already while reading the mesh on each MPI process.

The new class \texttt{parallel::fullydistributed::Triangulation} targets this issue
by distributing also the coarse grid. Such
a triangulation can be created by providing to each process a \texttt{Triangulation\-De\-scrip\-tion::Description} struct, containing
1) the relevant data to construct the local part of the coarse grid, 2) the
translation of the local coarse-cell IDs to globally unique IDs, 3) the hierarchy
of mesh refinements, and 4) the owner of the cells on the active mesh level as well
as on the multigrid levels. For the current release, triangulations is set up
this way cannot be adaptively refined after construction, which is planned to be
improved for the next release.

@@

The new fully distributed triangulation class supports 1D, 2D, and 3D meshes
including geometric multigrid hierarchies, periodic boundary conditions, and
hanging nodes.

@@

The class \texttt{VectorizedArray<Number>} is a key ingredient for the high
node-level performance of the matrix-free algorithms in deal.II~\cite{KronbichlerKormann2012, KronbichlerKormann2019}. It is a wrapper
class around a short vector of $n$ entries of type \texttt{Number} and maps
arithmetic operations to appropriate single-instruction/multiple-data (SIMD)
concepts by intrinsic functions.
The class \texttt{VectorizedArray} has been made more user-friendly in this release by making
it compatible with the STL algorithms found in the header \texttt{<algorithm>}.
The length of the  vector can now be queried by  \texttt{VectorizedArray::size()} and its underlying number type by \texttt{VectorizedArray::value\_type}.
Furthermore, the \texttt{VectorizedArray} class now supports range-based iterations over its entries.

Up to release 9.1, the
vector length $n$ has been set at compile time of the library to the highest
possible value supported by the given processor architecture.
Now, a second optional template argument
\texttt{VectorizedArray<Number, size>} can be given with \texttt{size} explicitly controlling
the vector length within the capabilities of a particular instruction set.
A full list of supported
vector lengths is presented in Table~\ref{tab:vectorizedarray}.

To account for the variable-size \texttt{VectorizedArray} class, all matrix-free related classes (like \texttt{MatrixFree} and \texttt{FEEvaluation})
have been extended with a new optional template argument specifying the
\texttt{VectorizedArrayType}. This allows users to select the vector length/ISA and,
as a consequence, the number of cells to be processed at once directly in their applications:
The deal.II-based
library \texttt{hyper.deal}~\cite{munch2020hyperdeal}, which solves the 6D Vlasov--Poisson equation with high-order
discontinuous Galerkin methods (with more than a thousand degrees of freedom per cell), constructs a tensor product of two \texttt{MatrixFree} objects of different SIMD-vector
length in the same application and benefits---in terms of performance---by the possibility of decreasing the number of cells processed by a single SIMD instruction.

\begin{table}
\caption{Supported vector lengths of the class \texttt{VectorizedArray} and
the corresponding instruction-set-architecture extensions. }\label{tab:vectorizedarray}
\centering
\begin{tabular}{ccc}
\toprule
\textbf{double} & \textbf{float} & \textbf{ISA}\\
\midrule
VectorizedArray<double, 1> & VectorizedArray<float, 1> & (auto-vectorization) \\
VectorizedArray<double, 2> & VectorizedArray<float, 4> & SSE2/AltiVec \\
VectorizedArray<double, 4> & VectorizedArray<float, 8> & AVX/AVX2 \\
VectorizedArray<double, 8> & VectorizedArray<float, 16> & AVX-512 \\
\bottomrule
\end{tabular}

\caption{Comparison of relevant SIMD-related classes in deal.II and \texttt{C++23}.}\label{tab:simd}
\centering
\begin{tabular}{cc}
\toprule
\textbf{VectorizedArray (deal.II)} & \textbf{std::simd (\texttt{C++23})} \\
\midrule
VectorizedArray<Number> & std::experimental::native\_simd<Number> \\
VectorizedArray<Number, size> & std::experimental::fixed\_size\_simd<Number, size> \\ \bottomrule
\end{tabular}
\end{table}

Furthermore, the new interfaces enable using any data structure
\texttt{VectorizedArrayType} as long as it supports required
functionalities like \texttt{size()} or \texttt{value\_type}. This prepares
for the \texttt{C++23} feature \texttt{std::simd} that will be enabled in the future.
Table~\ref{tab:simd} gives a comparison of the deal.II-specific SIMD classes and
the equivalent \texttt{C++23} classes. Finally, this change also prepares for specialized
code paths exploiting
vectorization within an element~\cite{KronbichlerKormann2019} in the future.

%We welcome the standardization of the SIMD
%parallelization paradigm in C++ and intend to replace step by step our own
%wrapper class, which has been continuously developed over the last decade. We
%would like to emphasize that the work invested in this class was not in vain,
%since many performance-relevant utility functions implemented with \texttt{VectorizedArray} in mind (e.g., \texttt{vectorized\_load\_and\_transpose}
%and \texttt{vectorized\_transpose\_and\_store}) will be still used, since they
%have not become part of the standard.

@@

@@

Large-scale simulations with 304,128 cores have revealed bottlenecks in release 
9.1 during initialization of a number of distributed data structures due to the usage of expensive collective operations 
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak
  Alltoall()}. Typical examples are the
pre-computation of those indices of the vector entries (or
other linear index ranges) owned by
each process, which were previously stored in  an array on every process. 
This information is needed to set up 
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class.
In release 9.2, we have replaced these functions in favor of
consensus algorithms~\cite{hoefler2010scalable}, which can be
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}).
Now, only the locally relevant information about the index ranges is
(re)computed when needed, which, for more than 100 MPI processes, uses
point-to-point communications and a single \texttt{MPI\_IBarrier()}.

@@

\todo[inline]{Aren't the target processes the result of the operation, and the
  input global numbers of indices that we request?}

By replacing the collective communications during set up and removing the arrays
that contain information for each process (enabled by the application of consensus
algorithms and other modifications---a full list of modifications leading to this
improvement can be found online), we were able to significantly
improve the set up
time for large-scale simulations and to solve a Poisson problem with multigrid
with \num{12e12} unknowns.
\todo[inline]{Where did we solve for 12T unknowns? The largest I [Martin] have done is 2.1T unknowns.}
Figure~\ref{} compares the timings of a simulation (including set up) with the
previous release 9.1 and with the current release 9.2.
{\color{red}TODO[Peter/Martin] description of the results}

The new code has been also applied to solve problems with adaptively refined
meshes with more than \num{4e9} unknowns. {\color{red}TODO[Timo]}

\todo[inline]{We did solve adaptive problems with more than 4B unknowns before, but some were broken for a while during this release stage due to the consensus algorithms. But we did not test systems or block vectors before. But we should not make it sound too negative.}

@@

With the previous release, \dealii supports $hp$-adaptive finite element methods
on distributed memory systems \cite{dealII91}. We implemented the bare functionality
for $hp$-adaptive methods with the objective to offer the greatest flexibility in
their application. Here, reference finite elements still had to be assigned manually
to each cell, which may not lead to an optimal mesh and is tedious.

With the current release, $hp$-adaptive finite element methods have been further
expanded: New features like decision strategies have been added and the user interface
has been overhauled, effectively making $hp$-methods more attractive to use. We introduced
many new functions that automatize the general workflow for applying $hp$-decision
strategies, which run on top of the previous low-level implementation for both serial
and parallel applications.

The interface is now as simple to use as the one for pure adaptive mesh refinement.
Consider the following (incomplete) listing as an example: We estimate both error and
smoothness of the finite element approximation. Further, we flag certain fractions of
cells with the highest and lowest errors for refinement and coarsening, respectively
(here: 30\% / 3\%). From those cells listed for adaptation, we specify a subset of them
for $h$- and $p$-adaptation (here: 10\% / 90\%).
\begin{c++}
Vector<float> estimated_error_per_cell (n_active_cells);
KellyErrorEstimator::estimate(
  dof_handler, ..., solution, estimated_error_per_cell, ...);
GridRefinement::refine_and_coarsen_fixed_number(
  triangulation, estimated_error_per_cell, 0.3, 0.03);

Vector<float> estimated_smoothness_per_cell (n_active_cells);
SmoothnessEstimator::Legendre::coefficient_decay(
  ..., dof_handler, solution, estimated_smoothness_per_cell);
hp::Refinement::p_adaptivity_fixed_number(
  dof_handler, estimated_smoothness_per_cell, 0.9, 0.9);
hp::Refinement::choose_p_over_h (dof_handler);

triangulation.execute_coarsening_and_refinement();
\end{c++}

In particular, we implemented decision strategies based on refinement history and
smoothness estimation, and made sure that they work for refinement as well as
coarsening in terms of $h$- and $p$-adaptation in serial and parallel applications.

The former relies on knowing an estimate for the upper error bound \cite[Thm.~3.4]{BabuskaSuri1990}.
For successive refinements, we can predict how the error will change on basis of
current error estimates and adaptation flags. In the next refinement cycle, these
predicted error estimates allow us to decide whether the choice of adaptation in
the previous cycle was justified, and provide the foundation for it in the next
one \cite{MelenkWohlmuth2001}.

In general, $p$-refinement is favorable over $h$-refinement in smooth regions of
the finite element approximation \cite[Thm.~3.4]{BabuskaSuri1990}. Thus, estimating
its smoothness provides a suitable decision indicator for $hp$-adaptation. For this
purpose, we express the finite element approximation in an orthogonal
basis of increasing frequency, and consider the decay of their expansion
coefficients as the estimation of smoothness. This has been implemented for both
Fourier coefficients \cite{BangerthKayserHerold2007} and Legendre coefficients
\cite{Mavriplis1994,HoustonSeniorSueli2003,HoustonSueli2005,EibnerMelenk2007}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Improved performance of symbolic framework}
\label{subsec:symbdiff}

\todo[inline]{J-P's section}

BatchOptimizer for symbolic expressions
\begin{itemize}
\item CSE for dictionary-based expressions
\item ``lambdify'' (via SymEngine)
\item Offloading to LLVM JIT compiler (via SymEngine)
\item Serialization, so complex expressions can be compiled offline
\end{itemize}

@@

\item \texttt{step-65} presents \texttt{TransfiniteInterpolationManifold}, a
manifold class that can propagate curved boundary information into the
interior of a computational domain by transfinite interpolation \cite{Gordon82}.
This manifold is a prototype for many other manifolds in that it is relatively
expensive to compute the new points, especially for higher order mappings. Since
typical programs query higher-order geometries in a large variety of contexts,
the contribution of the mapping to the run time can be significant. As a
solution, the tutorial also presents the class \texttt{MappingQCache}, which
samples the information of expensive manifolds in the points of a
\texttt{MappingQ} and caches it. The tutorial shows that this makes all queries
to the geometry very cheap.

\item \texttt{step-67} is an explicit time integrator for the
compressible Euler equations discretized with a high-order discontinuous
Galerkin scheme using the matrix-free infrastructure. Besides the use of
matrix-free evaluators for systems of equations and over-integration, it also
presents \texttt{MatrixFreeOperators::CellwiseInverseMassMatrix}, a fast implementation
of the action of the inverse mass matrix in the DG setting using tensor
products. Furthermore, this tutorial demonstrates the usage of new
pre and post operations, which can be passed to
\texttt{MatrixFree::cell\_loop()}, to schedule operations on sections of vectors close
to the matrix-vector product to increase data locality.

@@

\item \texttt{step-70}
  \todo[inline]{Luca, please write.}
\end{itemize}

@@

The original \texttt{\dealii} paper containing an overview of its
architecture is \cite{BangerthHartmannKanschat2007}. If you rely on
specific features of the library, please consider citing any of the
following:
\begin{itemize}
 \item For geometric multigrid: \cite{Kanschat2004,JanssenKanschat2011,ClevengerHeisterKanschatKronbichler2019};
 \item For distributed parallel computing: \cite{BangerthBursteddeHeisterKronbichler11};
 \item For $hp$~adaptivity: \cite{BangerthKayserHerold2007};
  \item For partition-of-unity (PUM) and enrichment methods of the
    finite element space: \cite{Davydov2016};
 \item For matrix-free and fast assembly techniques:
   \cite{KronbichlerKormann2012,KronbichlerKormann2019};
 \item For computations on lower-dimensional manifolds:
   \cite{DeSimoneHeltaiManigrasso2009};
 \item For curved geometry representations and manifolds:
   \cite{HeltaiBangerthKronbichlerMola2019};
 \item For integration with CAD files and tools:
   \cite{HeltaiMola2015};
 \item For Boundary Elements Computations:
   \cite{GiulianiMolaHeltai-2018-a};
 \item For \texttt{LinearOperator} and \texttt{PackagedOperation} facilities:
   \cite{MaierBardelloniHeltai-2016-a,MaierBardelloniHeltai-2016-b}.
 \item For uses of the \texttt{WorkStream} interface:
   \cite{TKB16};
   \item For uses of the \texttt{ParameterAcceptor} concept, the
     \texttt{MeshWorker::ScratchData} base class, and the
     \texttt{ParsedConvergenceTable} class: \cite{SartoriGiulianiBardelloni-2018-a}.
\end{itemize}

\dealii can interface with many other libraries:
\todo[inline]{We picked up gingko. Anything else?}
\begin{multicols}{3}
\begin{itemize}
\item ADOL-C \cite{Griewank1996a,adol-c}
\item ARPACK \cite{arpack}
\item Assimp \cite{assimp}
\item BLAS and LAPACK \cite{lapack}
\item cuSOLVER \cite{cusolver}
\item cuSPARSE \cite{cusparse}
\item Gmsh \cite{geuzaine2009gmsh}
\item GSL \cite{gsl2016}
\item Ginkgo \cite{ginkgo-web-page}
\item HDF5 \cite{hdf5}
\item METIS \cite{karypis1998fast}
\item MUMPS \cite{ADE00,MUMPS:1,MUMPS:2,mumps-web-page}
\item muparser \cite{muparser-web-page}
\item nanoflann \cite{nanoflann}
\item NetCDF \cite{rew1990netcdf}
\item OpenCASCADE \cite{opencascade-web-page}
\item p4est \cite{p4est}
\item PETSc \cite{petsc-user-ref,petsc-web-page}
\item ROL \cite{ridzal2014rapid}
\item ScaLAPACK \cite{slug}
\item SLEPc \cite{Hernandez:2005:SSF}
\item SUNDIALS \cite{sundials}
\item SymEngine \cite{symengine-web-page}
\item TBB \cite{Rei07}
\item Trilinos \cite{trilinos,trilinos-web-page}
\item UMFPACK \cite{umfpack}
\end{itemize}
\end{multicols}
Please consider citing the appropriate references if you use interfaces to these
libraries.

@@

@@

M.~Kronbichler was supported by the German
Research Foundation (DFG) under the project ``High-order discontinuous
Galerkin for the exa-scale'' (\mbox{ExaDG}) within the priority program ``Software
for Exascale Computing'' (SPPEXA) and the Bayerisches Kompetenznetzwerk
f\"ur Technisch-Wissen\-schaft\-li\-ches Hoch- und H\"ochstleistungsrechnen
(KONWIHR) in the context of the project
``Performance tuning of high-order discontinuous Galerkin solvers for
SuperMUC-NG''.
